Imputation of Incomplete Multilevel Data with R

### 2024-01-23 Manuscript/manuscript.qmd

There are different strategies that can be adopted in the imputation process that account for clustering: inclusion of a cluster indicator variable, performing a separate imputation process for each cluster, or performing a imputation for each cluster simultaneously by using an imputation method that accounts for clustering.(Stata: https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/) TODO: replace ref.

@@

<!-- I don't understand the following paragraph. What is meant with "is restricted in datasets..."? -->
<!-- Also, this info is also mentioned in ##### Clustering Inclusion -->
Regarding the restrictions imposed by the data, for instance, the use of cluster indicator variables is restricted in datasets where there are not many clusters and many observations per cluster (Graham, 2009). The last restriction is also required when imputations are performed on each cluster separately.  When this restriction cannot be achieved, one can use an imputation model that simultaneously imputes all clusters using a hierarchical model (Allison 2002).

@@

Here it is important to evaluate the assumptions imposed by the main model, for instance the cluster indicator strategy bias the estimator when the model is based on a hierarchical model (Taaljard,2008). Although an imputation strategy congenial with the main model is preferred, it is important to consider whether a complex imputation model is appropriate for the data, as a less complex imputation strategies may also lead to unbiased estimates in certain scenarios (Bailey 2020). For instance, in causal effect analysis, separately imputating data sets may lead to smaller bias when the size of the smaller exposure cluster is large, compared with an imputation model that includes exposure-confounder interactions. (Zhang,2023).
<!-- What is an exposure cluster? -->

Below we provide a imputation workflow that can be used in general to impute cluster data.

@@

### Assess Estimation Procedure Robustness to Missing Data
Before diving into imputation, make sure your estimation model can handle missing data. Sometimes, simpler methods like complete case analysis might be suffice, especially if your missingness is low (usually <5%). There are scenarios in which specific Maximum Likelihood (ML) estimation methods outperform Multiple Imputation (MI) methods, for instance when the response variable is the sole incomplete variable, mixed models demonstrate robustness to missing data under the Missing at Random (MAR) assumption and with a correct variance-covariance specification.\cite{Molenberghs_2007}.

<!-- What does it mean if an estimation model is not able to handle missing data? What do you mean with estimation model? -->
<!-- AFAIK, when only the outcome is missing, MI and a ML based method lead to similar results. The text appears to suggest that in that case ML based methods are better. -->

@@

Deductive imputation is also useful for getting values for level-1 variables from level-2 ones, like in Individual Participant Data (IPD). In some situations, you can derive missing information from metadata or protocols. For example, you could deduce missing test values for patients who have passed away. 
<!-- I don't think this is the best example. Usually, test scores after death are not of interest. The test score before death may be missing for a reason, but this should be derived based on clinical knowledge. The example is too vague (and I can't readily improve it.) -->

@@

On the other hand, the formula option is useful in specifying complex imputation models with polynomial terms or interactions and compared with the prediction matrix method do not requires the inclusion of additional terms as Just Another Variable (JAV) \cite{buuren2018a}($\S$ 6.4).

@@

<!-- What is the FAV variable? -->

It suggests using '2l.2stage.bin' for the FAV variable and '2l.2stage.norm' for the age variable. However, after inspecting the age density plot, we consider modifying its method to
'2l.2stage.pmm'. For the BMI variable, we employ deterministic imputation.

```{r obesitymeth, eval=TRUE}
meth_mar <-
  find.defaultMethod(
    Obesity,
    ind.clust = 1,
    I.small = 7,
    ni.small = 100,
    prop.small = 0.4
  )
meth_mar["BMI"] <- "~ I(Weight / (Height)^2)"
meth_mar["Age"] <- "2l.2stage.pmm"
```

<!-- What is suprime (below)? Do you mean suppress? -->

For these imputation models, it is necessary to specify the prediction matrix, with the cluster
variable labelled as -2 and the predictor variable measured within clusters labelled as 2, encompassing all variables. We need to suprime the variable Time as this variable is not specified in the main model. We also  modify the relationship between BMI, weight and height in the prediction matrix to avoid circular predictions. Then we proceed to run the imputation model.
```{r obesity-predmar,out.width="70%", eval=TRUE}
pred_mar <- quickpred(Obesity)
pred_mar[, "Cluster"] <- -2 
pred_mar[, "Time"] <- 0
pred_mar[pred_mar == 1] <- 2
pred_mar[c("Height", "Weight"), "BMI"] <- 0
plot_pred(pred_mar)
imp_mar <-
  mice(
    data = Obesity,
    method = meth_mar,
    predictorMatrix = pred_mar,
    printFlag = FALSE
  )
```

```{r obesity-predmar1,eval=TRUE}
summary(complete(imp_mar,"long")$Weight)
```
We also contemplate using the predictive mean matching (pmm) option, as the values imputed using a fully parametric method may be implausibly low for some patients.

```{r obesity-predmar_pmm, eval=TRUE}
meth_mar["Weight"] <- "2l.2stage.pmm"
imp_mar_pmm <- mice(
  data = Obesity,
  method = meth_mar,
  predictorMatrix = pred_mar,
  printFlag = FALSE
)
```

```{r obesity-predmar_pmm1,eval=TRUE}
summary(complete(imp_mar_pmm, "long")$Weight)
plot_trace(imp_mar_pmm, "Weight")
ggmice(imp_mar_pmm, aes(.imp, Weight)) + 
  geom_jitter()
```

Analyze the imputed data.
```{R}
#| prompt: true
#| warning: false
# fit <- with(
#     imp_mar_pmm,
#     lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
#     )
fit <- with(
    imp_mar_pmm,
    lmer(BMI ~ Age + FamOb + Gender + (1 + Age | Cluster))
    )
est <- testEstimates(as.mitml.result(fit), extra.pars = TRUE)
```

After confirming convergence, we proceed to save the results for future use.  We consider the possibility that patients may not have been selected randomly, which would then have led to a distribution for weight that does not reflect the weight in the population. It's likely that an omitted variable, like self-esteem, could influence this selection. For instance, individuals with lower self-esteem might have higher weight values, impacting their willingness to provide honest information due to embarrassment.

@@

```{r obesity-predmnar1, eval=TRUE}
meth_mnar <- meth_mar
meth_mnar["Weight"] <- "2l.2stage.heckman"
```
Then we proceed to run the imputation model as before, after executing these imputation procedures, it is essential to assess convergence and the coherence of the imputed values.
```{r obesity-predmnar2,eval=TRUE}
imp_mnar <- mice(
  data = Obesity,
  method = meth_mnar,
  predictorMatrix = pred_mnar,
  printFlag = FALSE
)
summary(complete(imp_mnar, "long")$Weight)
```

Upon examining the weight variable, we noticed that the imputed range falls outside the realm of plausible values, as weight should be positive.

```{r obesity-predmnar3,eval=TRUE}
summary(complete(imp_mnar, "long")$Weight)
```
Consequently, as before, we  use the 'pmm' option, but this time for the Heckman imputation. The pmm approach ensures that the imputed values remain within the range of observable values. This can be implemented by setting the pmm parameter to true.

```{r obesity-predmnarp, eval=TRUE}
imp_mnar_pmm <-
  mice(
    data = Obesity,
    method = meth_mnar,
    predictorMatrix = pred_mnar,
    pmm = T,
    printFlag = FALSE
  )
```

We check the convergence of the results
```{r obesity-predmnarp1, eval=TRUE}
summary(complete(imp_mnar_pmm, "long")$Weight)
plot_trace(imp_mar_pmm, "Weight")
```

After this modification we proceed to compare the effects on the model. We
run the analysis model on each of the completed datasets as well as the dataset where the
incomplete values are removed (Complete Case analysis, CC).

```{r models, eval=TRUE}
# cc_rs <- with(setDT(Obesity)[complete.cases(Obesity), ],
#               lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age |
#                     Cluster))
mar_rs <- with(
  imp_mar, 
  lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
  )
mar_pmm_rs <- with(
    imp_mar_pmm,
    lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
    )
mnar_rs <- with(
  imp_mnar, 
  lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
  )
mnar_pmm_rs <- with(
  imp_mnar_pmm,
  lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
  )
# list_models <- list(cc_rs, mar_rs, mar_pmm_rs, mnar_rs, mnar_pmm_rs)
# plot_models(list_models,
#             mod_name = c("Complete case", "MAR", "MAR_pmm", "MNAR", "MNAR_pmm"))
```

```{r}
est <- testEstimates(as.mitml.result(mnar_pmm_rs), extra.pars = TRUE)
```

```{r}
#| prompt: true
est$extra.pars |> 
  round(3) |>
  kable()
```
We note that there is minimal disparity in the age effect, FamObs, or  Gender across the various imputation models under consideration. An analysis of the intercept reveals that, under the MNAR assumption, a higher average BMI is anticipated compared to the MAR assumption.
Nonetheless, with respect to precision of the estimates, we notice that in general assuming MNAR leads to wider confidence intervals. In this case it does not have any influence on the final result but there could be cases where variation in the assumed missing mechanism could lead also to differences on the significance of a statistical test and therefore lead to contradictory conclusions.

>>>>

There are different strategies that can be adopted in the imputation process that account for clustering: inclusion of cluster indicator variable, performing a separate imputation process for each cluster, or performing a simultaneous imputation process by using an imputation method that accounts for clustering.\cite{zotero-516}

@@

<!-- I don't understand the following paragraph. What is meant with "is restricted in datasets..."?  I mean about the characteristics of the data such as sample size per cluster and balance of the data--> 
<!-- Also, this info is also mentioned in ##### Clustering Inclusion -->
Regarding the restrictions imposed by the nature of the data, i.e., the sample size, the use of cluster indicator variables is restricted in datasets where there are not many clusters and many observations per cluster (Graham, 2009). The last restriction is also required when imputations are performed on each cluster separately.  When this restriction cannot be achieved, one can use an imputation model that simultaneously imputes all clusters using a hierarchical model (Allison 2002).

@@

Here it is important to evaluate the assumptions imposed by the main model, for instance the cluster indicator strategy bias the estimator when the model is based on a hierarchical model (Taaljard,2008). Although an imputation strategy congenial with the main model is preferred, it is important to consider whether a complex imputation model is appropriate for the data, as a less complex imputation strategies may also lead to unbiased estimates in certain scenarios (Bailey 2020).
Below we provide a imputation workflow that can be used in general to impute cluster data.

@@

### Assess Estimation Procedure Robustness to Missing Data
Before diving into imputation, make sure your estimation model can handle missing data. Sometimes, simpler methods like complete case analysis might be suffice, especially if your missingness is low (usually <5%). There are scenarios in which specific Maximum Likelihood (ML) estimation methods outperform Multiple Imputation (MI) methods, for instance when the response variable is the sole incomplete variable, mixed models demonstrate robustness to missing data under the Missing at Random (MAR) assumption and with a correct variance-covariance specification.\cite{molenberghs2007}.

<!-- What does it mean if an estimation model is not able to handle missing data? there are models that implicitaly are able to manage the missingness like the FML where in the likelihood it is used the information of the completed observations and the incomplete ones.  What do you mean with estimation model? the main model you use to analyse the data-->
<!-- AFAIK, when only the outcome is missing, MI and a ML based method lead to similar results. The text appears to suggest that in that case ML based methods are better. what i understood is ML are better in that case but please check it.-->

@@

Deductive imputation is also useful for getting values for level-1 variables from level-2 ones, like in Individual Participant Data (IPD). For example, deductive imputation can be employed to infer missing test values for patients associated with a center unit that lacks a test procedure or machine, utilizing accessible metadata or protocol details.
<!-- I don't think this is the best example. Usually, test scores after death are not of interest. The test score before death may be missing for a reason, but this should be derived based on clinical knowledge. The example is too vague (and I can't readily improve it.)? is the new example better for you? -->

@@

On the other hand, the formula option is useful in specifying complex imputation models with polynomial terms or interactions and compared with the prediction matrix method do not requires the inclusion of additional terms as Just Another Variable (JAV) \cite{buur18}($\S$ 6.4).

@@

It suggests using '2l.2stage.bin' for the FamOb variable and '2l.2stage.norm' for the age variable. However, after inspecting the age density plot, we consider modifying its method to
'2l.2stage.pmm'. For the BMI variable, we employ deterministic imputation.

```{r obesitymeth, eval=TRUE}
meth_mar <-
  find.defaultMethod(
    Obesity,
    ind.clust = 1,
    I.small = 7,
    ni.small = 100,
    prop.small = 0.4
  )
meth_mar["BMI"] <- "~ I(Weight / (Height)^2)"
meth_mar["Age"] <- "2l.2stage.pmm"
meth_mar["Weight"] <- "2l.2stage.pmm"

```

For these imputation models, it is necessary to specify the prediction matrix, with the cluster
variable labelled as -2 and the predictor variable measured within clusters labelled as 2, encompassing all variables. We need to supress the variable Time as this variable is not specified in the main model. We also  modify the relationship between BMI, weight and height in the prediction matrix to avoid circular predictions. Then we proceed to run the imputation model.

```{r obesity-predmar,out.width="70%", eval=TRUE}
pred_mar <- quickpred(Obesity)
pred_mar["Weight",] <- 2
pred_mar["Weight","Weight"] <- 0
pred_mar[, "Cluster"] <- -2 
pred_mar[, "Time"] <- 0
pred_mar[pred_mar == 1] <- 2

pred_mar[c("Height", "Weight"), "BMI"] <- 0
#plot_pred(pred_mar)
imp_mar <-
  mice(
    data = Obesity,
    method = meth_mar,
    predictorMatrix = pred_mar,
    printFlag = FALSE
  )
```
After confirming convergence, we proceed to save the results for future use.

```{r obesity-predmar_pmm1,eval=TRUE}
summary(complete(imp_mar, "long")$Weight)
plot_trace(imp_mar, "Weight")
ggmice(imp_mar, aes(.imp, Weight)) + 
  geom_jitter()
```

We consider the possibility that patients may not have been selected randomly, which would then have led to a distribution for weight that does not reflect the weight in the population. It's likely that an omitted variable, like self-esteem, could influence this selection. For instance, individuals with lower self-esteem might have higher weight values, impacting their willingness to provide honest information due to embarrassment.

@@

```{r obesity-predmnar1, eval=TRUE}
meth_mnar <- meth_mar
meth_mnar["Weight"] <- "2l.2stage.heckman"
```

As before we  use the 'pmm', option but this time for the Heckman imputation, this approach ensures that the imputed values remain within the range of observable values. We then run the MNAR imputation model by setting the pmm parameter to true.

```{r obesity-predmnarp, eval=TRUE}
imp_mnar <-
  mice(
    data = Obesity,
    method = meth_mnar,
    predictorMatrix = pred_mnar,
    pmm = T,
    printFlag = FALSE
  )
```

We check the convergence of the results
```{r obesity-predmnarp1, eval=TRUE}
summary(complete(imp_mnar, "long")$Weight)
plot_trace(imp_mar, "Weight")
```

We proceed to compare the effects on the model. We
run the analysis model on each of the completed datasets as well as the dataset where the
incomplete values are removed (Complete Case analysis, CC).

```{r plotfunction, eval=TRUE, echo=FALSE}
library(broom.mixed)
plot_models <- function(list_models, mod_name){
  mod_table<-NULL
  for(i in 1:length(list_models)){
    mod <- list_models[[i]]  
    if(class(mod)[[1]]=="mira"){
      mod_tab <-summary(pool(mod), conf.int = TRUE)[,c("term","estimate","2.5 %","97.5 %","p.value")]
    }else{
      sum_cc <- (summary(mod,conf.int = TRUE))
      mod_tab<-data.frame(nlme::intervals(mod, which = "fixed")$fixed)
      mod_tab$term<-rownames(mod_tab)
      mod_tab$pvalue<-sum_cc$tTable[,5]
      mod_tab <-mod_tab[,c("term","est.","lower","upper","pvalue")]
      rownames(mod_tab)<-NULL

    }
    colnames(mod_tab)<-c("Coefficient","Estimate","lower","upper","p.value")
    mod_tab$model<-mod_name[i]
    mod_table<-rbind(mod_table,mod_tab)
  }

  plot <-ggplot(mod_table, aes(x = Coefficient, y = Estimate, ymin = lower, ymax = upper, group=model, color=model)) + 
    geom_hline(yintercept=0,linetype="dashed",color="black",size=0.1)+
    geom_linerange( position = position_dodge(width = 0.8)) + 
    geom_pointrange(size=0.1, position = position_dodge(width = 0.8))+coord_flip()+
    theme_classic()+theme(legend.position = "bottom")+scale_color_viridis_d()
  return(plot)
}
```
We proceed to fit the mixed model by using the complete case database and the imputed datasets (MAR,MNAR), and we compare the estimates of each dataset by using a plot (the plot function can be found in the appendix).
```{r models, eval=TRUE,warning=FALSE,message=FALSE}
# Save all datasets in a list
list_data <- list(Obesity |> filter(complete.cases(Obesity)), # complete case
                  imp_mar, # imputed under mar
                  imp_mnar) # imputed under mnar

# Run mixed models on each dataset
list_models <- lapply(list_data, 
                   FUN = function(x) 
                     with(x,nlme::lme(BMI ~ Age + FamOb + Gender, 
                                      random =  ~ 1 + Age | Cluster)))
# Plot coefficients
plot_models(list_models,
             mod_name = c("Complete case", "MAR", "MNAR"))
```

We note that there is minimal disparity in the age effect, FamObs, or  Gender across the various imputation models under consideration. An analysis of the intercept reveals that, under the MNAR assumption, a higher average BMI is anticipated compared to the MAR assumption.
Nonetheless, with respect to precision of the estimates, we notice that in general assuming MNAR leads to wider confidence intervals. In this case it does not have any influence on the final result but there could be cases where variation in the assumed missing mechanism could lead also to differences on the significance of a statistical test and therefore lead to contradictory conclusions.
